---
title: "Xenium_transcript_isolation_15dpf_32"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
 
```{r}
library(edgeR)
library(gplots)
library(plyr)
library(corrplot)
library(ggplot2)
library(EnhancedVolcano)
#install.packages('ggfortify')
library(ggfortify)
library(dplyr)
library(tibble)
```

```{r} 
Sample_info <- read.csv(file="Sample_info.csv", stringsAsFactors = FALSE)
(Sample_info)
```
#Combine the lists for each respective element from the sample info list
```{r}
AnnotInfo_all <- read.csv(file="featurecounts_readcounts_ann_2.csv",stringsAsFactors = FALSE,row.names=1)
AnnotInfo <- AnnotInfo_all[, -c(2, 3, 4, 5, 6, 17)]
head(AnnotInfo)
nodupAnnot <- AnnotInfo %>% distinct(Gene.Name, .keep_all = TRUE)
rownames(nodupAnnot) <- nodupAnnot[,1]
nodupAnnot<- nodupAnnot[,-1]
head(nodupAnnot)
table(duplicated(nodupAnnot[1]))#Still duplicates? actually these are in the rest of the table. The removal matches duplicate removal in excel.
which(duplicated(nodupAnnot[1]))
dim(AnnotInfo)
dim(nodupAnnot)
```
## Some exploratory analysis
```{r}
Normalized<-read.csv(file="normalized_data.csv",stringsAsFactors = FALSE,row.names = 1)
head(Normalized)
which(is.na(Normalized))
```
## Some exploratory analysis
```{r}
#raw<-read.csv(file="featdurecounts_readcounts_ann_2.csv",stringsAsFactors = FALSE,row.names = #1)
#head(raw)
#```


#R<-nodupAnnot
R<-nodupAnnot
head(R)
colnames(R)<-Sample_info$Sample
```

```{r}
#R<-raw[,-1]
#head(R)
#3colnames(R)<-Sample_info$Sample
```

#overall data summary
```{r}
summary(rowSums(R)) 
summary(rowMeans(R))
hist(log10(rowSums(R)), breaks = 10)
any(is.na(R))
sum(rowMeans(R, na.rm = TRUE) > 10)
sum(rowSums(R, na.rm = TRUE) > 10)

sum(rowMeans(R) > 10)
sum(rowSums(R) > 10)
boxplot(log2(abs(R)))
```
## Most highly expressed transcripts

```{r}
## Top 100
Top100<-rowSums(R)[order(rowSums(R), 
                                decreasing = TRUE)][1:100]
names(Top100)
R[names(Top100),]

```
## Top 10 transcript fraction of total counts
```{r,fig.height=4}
pdf(file="Top100FractionAll.pdf",height=4, width=6)
barplot(rowSums(R)[Top100] / sum(R), 
        ylab = "Count Fraction", 
        las = 2,
        #ylim = c(0, max(Top100)),
        #xlim = c(0, nrow(rowdata)),
        cex.names = 0.7)

dev.off()
```
## Lowly expressed transcripts
```{r}
# Total counts per transcript
TCounts<-rowSums(R)

# How many transcripts had 0 counts
length(which(TCounts==0))
```
```{r, fig.height=4}
hist(log2(TCounts))
abline(v = 2, col = "red", lty = 3, lwd = 2)
```
```{r}
## Only include transcripts with at least ~ 1 count per sample
Exp<-R[TCounts>1,]
dim(Exp)
```
```{r}
# Total counts per sample
SCounts<-colSums(Exp)
head(SCounts)
```
```{r,eval=FALSE}
#barplot(SCounts, las = 2)
   #     ylim = c(0, max(Top100)),
  #      xlim = c(0, count(rowdata[,1]))
safe_max <- function(x) {
 mx <- max(x, na.rm = TRUE)
  if (is.finite(mx)) {
    return(mx)
  } else {
    return(1)  # or any other default value you see fit
  }
}

#barplot(SCounts, las = 2, ylim = c(0, safe_max(Top100)), xlim = c(0, nrow(R)))
barplot(SCounts, las = 2)
```

Do a pca with all  
```{r}
Rawpc <- prcomp(t(R))

plot(Rawpc$x[,1], Rawpc$x[,2], col=as.factor(Sample_info$Location), pch=16)
text(Rawpc$x[,1], Rawpc$x[,2], labels=colnames(R), pos=1, cex=0.8)
```
Dendrogram
```{r}
Rawsd <- apply(R,2,sd)
hist(Rawsd)
Raw_bigsd <-R[which(Rawsd > quantile(Rawsd,.75)),]
dim(Raw_bigsd)
Rawdists <- dist(t(R))
Rawclust <- hclust(Rawdists)
plot(Rawclust, labels=Sample_info$SampleID)
```
## Statistical analysis with EdgeR  
Make DGE object and filter low expression genes  
```{r}
DGE1 <- DGEList(counts = R, genes=R, group=Sample_info$Location)
DGE1$samples
dim(DGE1)
keep <- filterByExpr(DGE1)
table(keep)
DGE1 <- DGE1[keep, , keep.lib.sizes=FALSE]
```
How many genes expressed above threshold?  
```{r}
dim(DGE1)
```
Normalize and calculate dispersion 
```{r}
DGE1 <- calcNormFactors(DGE1)
DGE1 <- estimateDisp(DGE1)
plotMDS(DGE1)
```
```{r}
d

```


```{r}
#result is exact test p value comparing GU and GI with GU reference
DGE1$samples$group <- relevel(DGE1$samples$group, ref="GU")
As_ET<-exactTest(DGE1,pair=c("GU","GI"))
Res_ET<-topTags(As_ET,n=Inf,sort.by="none")$table
Res_ET
```
Export csv with all genes
```{r}
write.csv(Res_ET,file="Bladder_development_output_all.csv")
```
### Write table for background (all gene symbols in dataset)
```{r}
write.table(Res_ET$Gene.Name,col.names=FALSE,row.names=FALSE,quote=FALSE,file="Background_Symbols.txt")
```

Export csv with all genes
```{r}
write.csv(Res_ET,file="Bladder_Development_gene_Table_all.csv")
```

```{r}
DevFDR <- Res_ET[which(Res_ET$FDR < .05 & abs(Res_ET$logFC)>1),]
```
```{r}
DevFDR1 <- Res_ET[which(Res_ET$FDR < .1 & abs(Res_ET$logFC)>1),]
```

```{r}
DevNomSig <- Res_ET[which(Res_ET$PValue < .05 & abs(Res_ET$logFC)>1),]
```
```{r}
Sig_Up<-Res_ET[which(Res_ET$PValue < .05 & Res_ET$logFC>1),]
dim(Sig_Up)
```
```{r}
write.table(Sig_Up$Gene_Name,col.names=FALSE,row.names=FALSE,quote=FALSE,file="Sig_Up_Symbols.txt")
```
```{r}
Sig_Dn<-Res_ET[which(Res_ET$PValue < .05 & Res_ET$logFC<(-1)),]
dim(Sig_Dn)
```
```{r}
write.table(Sig_Dn$Gene_Name,col.names=FALSE,row.names=FALSE,quote=FALSE,file="Sig_Dn_Symbols.txt")
```

```{r}
DevNoSig <- Res_ET[which(Res_ET$PValue < .05),]
```

Export csv with all genes
```{r}
write.csv(DevNomSig,file="Bladder_DevNomSig.csv")
```

Number of genes up and down
```{r}
length(which(DevNomSig$logFC>0))
length(which(DevNomSig$logFC<0))
```
Get top 50 Dev genes
```{r}
Devtop50genes<-DevNomSig[order(abs(DevNomSig$logFC),decreasing=TRUE),][1:50,"Gene_Name"]
Devtop50genes
```

Smear plot for tags < 0.05  
```{r}
plotSmear(DGE1, de.tags=rownames(DevNomSig), main = "Developmental Effect p-value < 0.05",ylab="Log2 Fold Change",cex=.75)
abline(h = c(-1, 1), col = "dodgerblue")
pdf("DevEffect_smearplot.pdf", height=6, width=9)
plotSmear(DGE1, de.tags=rownames(DevNomSig), main = "Developmental Effect p-value < 0.05",ylab="Log2 Fold Change",cex=.75)
abline(h = c(-1, 1), col = "dodgerblue")
dev.off()
```

Volcano  
```{r}
pdf("DevSig_volcanoplot.pdf", height=4, width=4)
par(mar=c(5,5,2,1))
plot(-log10(Res_ET$FDR) ~ Res_ET$logFC, main="", ylab="Adjusted P Value (-log10)", xlab="Log2 Fold Change",pch=20, ylim=c(0,4))
points(DevFDR1$logFC,-log10(DevFDR1$FDR),pch=20,col="blue")
abline(h=-log10(0.1))
abline(v=-1)
abline(v=1)
dev.off()
```

Fancier volcano
```{r}
Res_Volcano<-Res_ET[,c("Gene_Name","logFC","FDR")]
pdf("DevSig_enhvolcano.pdf", height=9, width=9)
EnhancedVolcano(Res_Volcano,lab=Res_Volcano$Gene_Name,x="logFC",y="FDR",
    pCutoff = .1,
    FCcutoff = 1,
    pointSize = 3.0,
    labSize = 5.0,
    ylim = c(0, 4),
    )
dev.off()
```



### Make a ranked gene list file for GSEA

First assigning EdgeRdata to a new object and making new columns to calculate score for GSEA
```{r}
GSEAdata<-Res_ET
GSEAdata$FCsign<-sign(GSEAdata$logFC)
GSEAdata$logP<- -log10(GSEAdata$PValue)
GSEAdata$GSEA_Score<-GSEAdata$logP/GSEAdata$FCsign
```
Sort the data by GSEA_Score
```{r}
GSEAdata<-GSEAdata[order(GSEAdata$GSEA_Score,decreasing=TRUE),]
head(GSEAdata)
```

Write a .rnk file with gene symbols and their scores

```{r}
write.table(GSEAdata[,c("Gene_Name","GSEA_Score")],sep="\t",col.names=FALSE,row.names=FALSE,quote=FALSE,file="Scores.rnk")
```

Read in GFGR pathway genes
```{r}
FGFRgenes<-read.csv("FGFR1genes.csv",header=FALSE)
```

```{r}
FGFRgeneExp<-Res_ET[Res_ET$Gene_Name%in%FGFRgenes$V1,]
write.csv(FGFRgeneExp,file="FGFR1pathway_BladderExp.csv")
```

